### 2022-12-14 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@88e24362304b6916bc5f1dd8a2e5a49d3b4d0d7c 🚀

Understanding the ecological constraints that shape species interactions
networks and predicting their emerging structure are thus two
complementary aims of network ecology. This distinction between
understanding and predicting is essential when using statistical and
mathematical models in network ecology and interpreting them. On one
hand, null models help us identify potential ecological mechanisms that
drive species interactions and constrain ecological networks. Null
models generate a distribution for a target measure using a set of rules
that exclude the mechanism of interest (Fortuna and Bascompte 2006;
Delmas et al. 2019). The deviation between the model and empirical data
helps us evaluate the effect of this ecological process in nature. On
the other hand, predictive models can help fill many gaps on species
interactions data. A variety of such models have recently been developed
using machine learning and other statistical tools, most of which are
presented in Strydom et al. (2021). However, given the constrained
nature of ecological networks, it is surprising that the principle of
maximum entropy, a rigorous mathematical method designed for both the
analysis and prediction of constrained systems, has been barely used in
network ecology.

@@

Food-web properties that can be derived using MaxEnt are varied and
pertain to different elements of the network (i.e.~at the species
(node), the interaction (edge) or the community (network) levels).
Because MaxEnt is a method of finding least-biased probability
distributions given partial knowledge about a system, these properties
need to be represented probabilistically. For example, at the species
level, MaxEnt can be used to predict the distribution of trophic levels
among species, as well as the distribution of species' vulnerability
(number of predators) and generality (number of prey). By contrast, at
the interaction level, predictions can be made on the distribution of
interaction strengths in weighted food webs. At the community level, it
can generate probability distributions of many measures of their
emerging structure and of networks themselves (i.e.~a probability
distribution that specific network configurations are realized given the
model and constraints). Overall, the potential of this method in the
study of food webs is broad. The applicability and performance of MaxEnt
mostly depend on the ecological information available and on our
capacity to find the right set of state variables that best represent
natural systems and to translate them into appropriate statistical
constraints. Having a validated maximum entropy model for the system at
hand allows us to make least-biased predictions using a minimal amount
of data, as well as identify the most important ecological processes
shaping that system. In other words, MaxEnt is a valuable and
underexploited tool in network ecology that allows us to better
understand and predict the structure of ecological networks.

@@

The principle of maximum entropy is a robust mathematical method of
finding least-biased probability distributions that have some specified
properties given by prior knowledge about a system. We first applied
this conventional MaxEnt approach on food webs to predict species level
properties, namely the joint degree distribution and the degree
distribution of maximum entropy given known numbers of species and
interactions. We found that the joint degree distributions of maximum
entropy had a similar shape to the ones of empirical food webs in
high-connectance systems. However, these MaxEnt distributions were more
symmetric than the ones of empirical food webs when connectance was low,
which suggests that other constraints might be needed to improve these
predictions in low-connectance systems. Then, we used a slightly
different approach that aimed at finding heuristically the network
configuration with the highest SVD entropy, i.e.~whose vector of
relative singular values has maximum entropy. This network of maximum
entropy is the most complex, or random, given the specified structure.
We found that the heuristic maximum entropy model based on connectance
did not predict the structure of sampled food webs very well. However,
the heuristic maximum entropy model based on the entire joint degree
sequence, i.e.~on the number of prey and predators for each species,
gave more convincing results. Indeed, this model reproduced food-web
structure surprisingly well, including the highly informative motifs
profile. Nevertheless, it was not able to predict realistic food-chain
lengths.

>>>>

Understanding the ecological constraints that shape species interactions
networks and predicting their emerging structure are thus two
complementary aims of network ecology. This distinction between
understanding and predicting is essential when using statistical and
mathematical models in network ecology and interpreting them. On one
hand, null models help us identify potential ecological mechanisms that
drive species interactions and constrain ecological networks. Null
models generate a distribution for a target measure using a set of rules
that exclude the mechanism of interest (Fortuna and Bascompte 2006;
Delmas et al. 2019). The deviation between the model and empirical data
helps us evaluate the effect of this ecological process in nature. On
the other hand, predictive models can help fill many gaps on species
interactions data. A variety of such models have recently been developed
using machine learning and other statistical tools, most of which are
presented in Strydom et al. (2021). However, given the constrained
nature of ecological networks, it is surprising that the principle of
maximum entropy, a mathematical method designed for both the analysis
and prediction of constrained systems, has been barely used in network
ecology.

@@

Food-web properties that can be derived using MaxEnt are varied and
pertain to different elements of the network (i.e.~at the species
(node), the interaction (edge) or the community (network) levels).
Because MaxEnt is a method of finding least-biased probability
distributions given partial knowledge about a system, these properties
need to be represented probabilistically. For example, at the species
level, MaxEnt can be used to predict the distribution of trophic levels
among species, as well as the distribution of species' vulnerability
(number of predators) and generality (number of prey). By contrast, at
the interaction level, predictions can be made on the distribution of
interaction strengths in weighted food webs. At the community level, it
can generate probability distributions of many measures of their
emerging structure and of networks themselves (i.e.~a probability
distribution that specific network configurations are realized given the
model and constraints). Overall, the potential of this method in the
study of food webs is broad. The applicability and performance of MaxEnt
mostly depend on the ecological information available and on our
capacity to find the right set of state variables that best represent
natural systems and to translate them into appropriate statistical
constraints. Having a validated maximum entropy model for the system at
hand allows us to make least-biased predictions using a minimal amount
of data, as well as identify the most important ecological processes
shaping that system. In other words, MaxEnt can help us better
understand and predict the structure of ecological networks worldwide.

@@

The principle of maximum entropy is a mathematical method of finding
least-biased probability distributions that have some specified
properties given by prior knowledge about a system. We first applied
this conventional MaxEnt approach on food webs to predict species level
properties, namely the joint degree distribution and the degree
distribution of maximum entropy given known numbers of species and
interactions. We found that the joint degree distributions of maximum
entropy had a similar shape to the ones of empirical food webs in
high-connectance systems. However, these MaxEnt distributions were more
symmetric than the ones of empirical food webs when connectance was low,
which suggests that other constraints might be needed to improve these
predictions in low-connectance systems. Then, we used a slightly
different approach that aimed at finding heuristically the network
configuration with the highest SVD entropy, i.e.~whose vector of
relative singular values has maximum entropy. This network of maximum
entropy is the most complex, or random, given the specified structure.
We found that the heuristic maximum entropy model based on connectance
did not predict the structure of sampled food webs very well. However,
the heuristic maximum entropy model based on the entire joint degree
sequence, i.e.~on the number of prey and predators for each species,
gave more convincing results. Indeed, this model reproduced food-web
structure surprisingly well, including the highly informative motifs
profile. Nevertheless, it was not able to predict realistic food-chain
lengths.
